---
title: "Forest Removal"
author: "David Clark"
date: "July 3, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd('/Users/dave/DataScience/Projects/GitHub/Forest-Removal')
```

In this project I investigate the relationship between forest removal and
percent forest coverage for countries in the world. Forest removal is defined as
the total value in US dollars of all products removed during that year. Products
are both wood and non-wood related. All data is taken from the [Gapminder](https://www.gapminder.org/) website.

## Data Processing and Exploration

Let's load the data and inspect it.

```{r}
forestCoverage = read.csv("data/forest_coverage.csv", header = TRUE)
forestRemoval = read.csv("data/forest_products_per_ha.csv", header = TRUE, na.strings = "")
```

Let's view the column names:

```{r}
names(forestCoverage)
names(forestRemoval)
```

I am only interested in the columns with data containing the country name and 
data for a specific year. Therefore, I am interested in columns 1 - 4 for 
`forerstCoverage` and columns 1 - 2 for `forestRemoval`.

```{r}
forestCoverage = forestCoverage[1:4]
forestRemoval = forestRemoval[1:2]
```

Let's rename the columns so that they are more readable:

```{r}
colsForestCoverage = c("Country", "fc_1990", "fc_2000", "fc_2005")
colnames(forestCoverage) = colsForestCoverage

colsForestRemoval = c("Country", "fr_2005")
colnames(forestRemoval) = colsForestRemoval
```

In the `forestRemoval` data frame, the values in the `fr_2005` column are factors.
I need to convert them to numeric, so that I can perform a linear regression fit:

```{r}
forestRemoval$fr_2005 = as.numeric(levels(forestRemoval$fr_2005))[forestRemoval$fr_2005]
```

Now, let's view some metrics using `summary()`:

```{r}
summary(forestCoverage)
summary(forestRemoval)
```

You will notice that there are many points that are NAs These are missing data
values and I want to remove them from the analysis. I do this with the following
commands:

```{r}
forestCoverage = na.omit(forestCoverage)
forestRemoval = na.omit(forestRemoval)
```

Let's look at a summary of the data again, but this time with out the NA values.

```{r}
summary(forestCoverage)
summary(forestRemoval)
```

Since I am interested in the relationship between forest removal and forest
coverage, I need to compare each dataframe equally. This means, I am only
interested in countries that have data for both forest removal **and** forest
coverage. To ensure this, I will merge the two dataframes into one dataframe 
using country to connect the two.

```{r}
forestMerged = merge(forestCoverage, forestRemoval, by='Country')
```

Now I attach the columns to the dataframe so I can reference the columns by
column name:

```{r}
attach(forestMerged)
```

Before modeling the data, let's use `ggplot` to make a scatterplot of forest
removal versus forest coverage for the year 2005. This is the only year where
data exists for both set's of measurements.

```{r}
library(ggplot2)
```

```{r}
ggplot(forestMerged, aes(x=fc_2005, y=fr_2005)) +
        labs(title = 'Forest Removal', x='Forest Coverage (%)', y='Forest Removal') +
        geom_point()
```

Right off you will notice a wide scatter.  Specifically, there is a larger scatter
in forest removal for low values of forest coverage.

## Linear Regression

Now I am ready to start modeling the data.

First, I fit a linear regression model to this dataset to see what is the 
relationship between forest coverage and forest removal.

```{r}
lm.fit = lm(fr_2005 ~ fc_2005, data=forestMerged)

coef(lm.fit)
```

The negative coefficient for `fc_2005` indicates that there is an inverse
relationship between forest coverage and forest removal.

Let's look at the fit in more detail.

```{r}
summary(lm.fit)
```

Here the small R-squared value and an F-statistic with a value greater than one
indicate a reasonable model fit.

Can the model be improved by using a non-linear fit? Let's take the log of
forest removal and see how things change:

```{r}
lm.fit = lm(log(fr_2005) ~ fc_2005, data=forestMerged)
summary(lm.fit)
```

Now F-statistic and R-squared values have improved by almost a factor of two.
Clearly adding a log term has improved the model fit.

To investigate this model fit further, let's compute confidense intervals. This
is done using the function `confint()`:

```{r}
confint(lm.fit)
```

We can also compute the confidense interval at several points using the 
`predict()` function. I choose the percent forest coverage values of 20, 40, 60,
and 80:

```{r}
predict(lm.fit, data.frame(fc_2005=c(20, 40, 60, 80)), interval = 'confidence')
```

As expected, the fit decreases for larger values of forest coverage. You will
also notice that the confidense interval ranges in width from ~70 to 140. This
could be because the are fewer points for larger values of forest coverage, 
which makes it harder for the model to make a good fit.

I now replot the scatter plot and add the regression line. I use `geom_smooth()`
to add the regression line and include the confidense region on either side of
the line.

```{r}
ggplot(forestMerged, aes(x=fc_2005, y=fr_2005)) +
        labs(title = 'Forest Removal', x='Forest Coverage (%)', y='Forest Removal') +
        geom_point() +
        geom_smooth(method = lm)
```

Now we can graphically see that there is a downward trend in the model fit. This
indicates that countries with more forest tend to remove less wood. But, why does
this trend exist? Are there other factors that contribute to forest removal? To
investigate, let's add additional features and see how this affects the model
fit.

## Adding Features

One reason countries with more forest might remove less wood is because they have
a smaller population. Let's add population as an additional feature and see what 
we find.

Load the data:
```{r}
population = read.csv('data/population.csv', header = TRUE)
attach(population)
```

Since I am only modeling data from 2005, I am only interested in the population
for that year:

```{r}
population = population[c('Total.population', 'X2005')]
```

Rename the columns:

```{r}
colsPopulation = c("Country", "p_2005")
colnames(population) = colsPopulation
```

Merge population dataframe with `forestMerged` dataframe:

```{r}
forestPopMerged = merge(forestMerged, population, by="Country")
attach(forestPopMerged)
```

Performing a `summary()` on the `forestPopMerged` dataframe, we see that `p_2005`
is of type other. We need to set this to a numeric value:

```{r}
forestPopMerged$p_2005 = as.numeric(levels(forestPopMerged$p_2005))[forestPopMerged$p_2005]
summary(forestPopMerged)
```

Remove NA values:

```{r}
forestPopMerged = na.omit(forestPopMerged)
summary(forestPopMerged)
```

Due to the merge, some points in the `forestMerged` dataframe have been removed.
Let's redo the model fit and see how it has changed:

```{r}
lm.fit = lm(fr_2005 ~ fc_2005, data=forestPopMerged)
summary(lm.fit)
```

And `log(fr_2005)`...

```{r}
lm.fit = lm(log(fr_2005) ~ fc_2005, data=forestPopMerged)
summary(lm.fit)
```

Yes, it did improve!

Let's take a look at the plot.

```{r}
ggplot(forestPopMerged, aes(x=fc_2005, y=fr_2005)) +
        labs(title = 'Forest Removal', x='Forest Coverage (%)', y='Forest Removal') +
        geom_point() +
        geom_smooth(method = 'lm')
```

### Add population feature

Now, let's add population as an additional feature:

```{r}
lm.fit2 = lm(fr_2005 ~ fc_2005*p_2005, data=forestPopMerged)
summary(lm.fit2)
```

The smaller F-statistic and larger value of R-squared indicates that adding
population as an additional feature did not improve the model fit.

What if we take the log of population? Will this improve the fit?

```{r}
lm.fit2 = lm(log(fr_2005) ~ fc_2005*p_2005, data=forestPopMerged)
summary(lm.fit2)
```

No, it did not improve the model. The fit is worse.

What about `log(fr_2005)` as we did with the initial model?

```{r}
lm.fit2 = lm(log(fr_2005) ~ fc_2005*p_2005, data=forestPopMerged)
summary(lm.fit2)
```

No, not so great either. The best fit so far is with the log of forest removal
and forest coverage. I will continue to explore more features and see what other
factors contribute to forest removal.

