---
title: "Forest Removal"
author: "David Clark"
date: "July 3, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd('/Users/dave/DataScience/Projects/GitHub/Forest-Removal')
```

In this project I investigate the relationship between forest removal and
percent forest coverage for countries in the world. Forest removal is defined as
the total value in US dollars of all products removed during that year. Products
are both wood and non-wood related. All data is taken from the [Gapminder](https://www.gapminder.org/) website.

## Data Processing and Exploration

Let's load the data and inspect it.

```{r}
forestCoverage = read.csv("data/forest_coverage.csv", header = TRUE)
forestRemoval = read.csv("data/forest_products_per_ha.csv", header = TRUE, na.strings = "")
```

Let's view the column names:

```{r}
names(forestCoverage)
names(forestRemoval)
```

I am only interested in the columns with data containing the country name and 
data for a specific year. Therefore, I am interested in columns 1 - 4 for 
`forerstCoverage` and columns 1 - 2 for `forestRemoval`.

```{r}
forestCoverage = forestCoverage[1:4]
forestRemoval = forestRemoval[1:2]
```

Let's rename the columns so that they are more readable:

```{r}
colsForestCoverage = c("Country", "fc_1990", "fc_2000", "fc_2005")
colnames(forestCoverage) = colsForestCoverage

colsForestRemoval = c("Country", "fr_2005")
colnames(forestRemoval) = colsForestRemoval
```

In the `forestRemoval` data frame, the values in the `fr_2005` column are factors.
I need to convert them to numeric, so that I can perform a linear regression fit:

```{r}
forestRemoval$fr_2005 = as.numeric(levels(forestRemoval$fr_2005))[forestRemoval$fr_2005]
```

Now, let's view some metrics using `summary()`:

```{r}
summary(forestCoverage)
summary(forestRemoval)
```

You will notice that there are many points that are NAs These are missing data
values and I want to remove them from the analysis. I do this with the following
commands:

```{r}
forestCoverage = na.omit(forestCoverage)
forestRemoval = na.omit(forestRemoval)
```

Let's look at a summary of the data again, but this time with out the NA values.

```{r}
summary(forestCoverage)
summary(forestRemoval)
```

Since I am interested in the relationship between forest removal and forest
coverage, I need to compare each dataframe equally. This means, I am only
interested in countries that have data for both forest removal **and** forest
coverage. To ensure this, I will merge the two dataframes into one dataframe 
using country to connect the two.

```{r}
forestMerged = merge(forestCoverage, forestRemoval, by='Country')
```

Now I attach the columns to the dataframe so I can reference the columns by
column name:

```{r}
attach(forestMerged)
```

Before modeling the data, let's use `ggplot` to make a scatterplot of forest
removal versus forest coverage for the year 2005. This is the only year where
data exists for both set's of measurements.

```{r}
library(ggplot2)
```

```{r}
ggplot(forestMerged, aes(x=fc_2005, y=fr_2005)) +
        labs(title = 'Forest Removal', x='Forest Coverage (%)', y='Forest Removal') +
        geom_point()
```

Right off you will notice a wide scatter.  Specifically, there is a larger scatter
in forest removal for low values of forest coverage.

## Linear Regression

Now I am ready to start modeling the data.

First, I fit a linear regression model to this dataset to see what is the 
relationship between forest coverage and forest removal.

```{r}
lm.fit = lm(fr_2005 ~ fc_2005, data=forestMerged)

coef(lm.fit)
```

The negative coefficient for `fc_2005` indicates that there is an inverse
relationship between forest coverage and forest removal. In other words, 
countries with more forest tend to remove less material for products.

Let's look at the fit in more detail.

```{r}
summary(lm.fit)
```

Here the small R-squared value and an F-statistic with a value greater than one
indicate a reasonable model fit.

To investigate this model fit further, let's compute confidense intervals. This
is done using the function `confint()`:

```{r}
confint(lm.fit)
```

We can also compute the confidense interval at several points using the 
`predict()` function. I choose the percent forest coverage values of 20, 40, 60,
and 80:

```{r}
predict(lm.fit, data.frame(fc_2005=c(20, 40, 60, 80)), interval = 'confidence')
```

As expected, the fit decreases for larger values of forest coverage. You will
also notice that the confidense interval ranges in width from ~70 to 140. This
could be because the are fewer points for larger values of forest coverage, 
which makes it harder for the model to make a good fit.

I now replot the scatter plot and add the regression line. I use `geom_smooth()`
to add the regression line and include the confidense region on either side of
the line.

```{r}
ggplot(forestMerged, aes(x=fc_2005, y=fr_2005)) +
        labs(title = 'Forest Removal', x='Forest Coverage (%)', y='Forest Removal') +
        geom_point() +
        geom_smooth(method = lm)
```



